---
title: "Movielense Data: Collaborative Recommendation"
author: 
  - Noam Mansovsky, noamm2@illinois.edu
  - Tatiana Sokolinski, tatiana9@illinois.edu
date: "12/5/2021"
output:
  html_document: 
    theme: yeti
    toc: yes
urlcolor: cyan
---
***
```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen = 1, digits = 4, width = 80, fig.align = "center")

mypackages = c( "dplyr","ggplot2","recommenderlab","DT","data.table","reshape2","Matrix","proxy","doBy")   
tmp = setdiff(mypackages, rownames(installed.packages())) 
if (length(tmp) > 0) install.packages(tmp)

library(dplyr)
library(ggplot2)
library(Matrix)
library(recommenderlab)
library(DT)
library(data.table)
library(reshape2)

library(proxy)
library(doBy) # to get top-n elements
library(knitr)
```


## Read in Data

### Ratings data:
```{r}
myurl = "https://liangfgithub.github.io/MovieData/"
# ratings_data_location = paste0(myurl, 'ratings.dat?raw=true')
ratings_data_location =  'ratings.dat'
ratings = read.csv(ratings_data_location, 
                   sep = ':',
                   colClasses = c('integer', 'NULL'), 
                   header = FALSE)
colnames(ratings) = c('UserID', 'MovieID', 'Rating', 'Timestamp')
head(ratings, 5)
```


### User data:
```{r}
users = read.csv(paste0(myurl, 'users.dat?raw=true'),
                 sep = ':', header = FALSE)
users = users[, -c(2,4,6,8)] # skip columns
colnames(users) = c('UserID', 'Gender', 'Age', 'Occupation', 'Zip-code')
head(users, 5)
```

### Movies data:

```{r}
movies = readLines(paste0(myurl, 'movies.dat?raw=true'))
movies = strsplit(movies, split = "::", fixed = TRUE, useBytes = TRUE)
movies = matrix(unlist(movies), ncol = 3, byrow = TRUE)
movies = data.frame(movies, stringsAsFactors = FALSE)
colnames(movies) = c('MovieID', 'Title', 'Genres')
movies$MovieID = as.integer(movies$MovieID)

# convert accented characters
movies$Title[73]
movies$Title = iconv(movies$Title, "latin1", "UTF-8")
movies$Title[73]

# extract year
movies$Year = as.numeric(unlist(
  lapply(movies$Title, function(x) substr(x, nchar(x)-4, nchar(x)-1))))
head(movies, 5)
```

***
# System II: Collaborative Recommendation System

## 1. User-based recommendation:

User-Based Collaborative Filtering is an algorithm used to predict the items that a user might like on the basis of ratings given to that item by the other users who have similar taste with that of the target user.

- **Create the user profile matrix:**

Create ratings matrix using MovieID, UserID and Ratings fields.

```{r}
# Create ratings matrix. Rows = UserID, Columns = MovieID
ratingmat <- dcast(ratings, UserID~MovieID, value.var = "Rating", na.rm=FALSE)
ratingmat <- as.matrix(ratingmat[,-1]) #remove userIds
# Convert rating matrix into a recommenderlab sparse matrix
ratingmat <- as(ratingmat, "realRatingMatrix")
dim(ratingmat)
```

- **Train and test splits:**

The `recommenderlab::evaluationScheme()` function can split the data set into training and testing subsets:

- **train** matrix contains about 80% rows of `rating` dataset

- **test** matrix contains about 20% rows of `rating` dataset


The output of the evaluationScheme() function is a single R object containing both the training and testing subsets.

```{r}
set.seed(9335)
# Parameters:
percent_train = 0.8 # split the data into the training and the test sets (80%/20%)
items_to_keep = 15  # items to use for each user
rating_threshold = 4 # threshold at which ratings are considered good for evaluation
eval_sets <- evaluationScheme(ratingmat, method="split", train=percent_train, given=items_to_keep, goodRating=rating_threshold)
eval_sets
```

- **Build Recommender Model**

Build Recommender Model using User-Based Collaborative Filtering using `recommenderlab` library. Set method equals to "UBCF", use Cosine Similarity, Z-score normalization and 30 Nearest Neighbors as parameters.
```{r}
# Create Recommender Model using User-Based Collaborative Filtering
UBCF_model <- Recommender(getData(eval_sets, "train"), "UBCF", 
                           param=list(normalize = "Z-score",method="Cosine",nn=30))
```

```{r echo = FALSE, message = FALSE, warning = FALSE}
# Save the model
saveRDS(UBCF_model, file = "data/UBCF_recommender.rds")
# UBCF_model <- readRDS("data/UBCF_recommender.rds")
```

- **Evaluate your model:**

To evaluate the model we use the `recommenderlab` package. We first use the `predict()` function to generate predictions for the known portion of the test data. Then, the `calcPredictAccuracy()` function to calculate the error between the predictions and the unknown portions of the test data.

```{r}
items_to_recommend = 10
UBCF_eval_prediction = predict(object = UBCF_model,
                          newdata = getData(eval_sets, "known"),
                          n = items_to_recommend,
                          type = "ratings")
# Get model performance
UBCF_eval_accuracy = calcPredictionAccuracy(x = UBCF_eval_prediction,
                                       data = getData(eval_sets, "unknown"),
                                       byUser = FALSE)
UBCF_eval_accuracy
```





- **Get recommendations:**


```{r}
# Obtain top 10 recommendations for n user in dataset
items_to_recommend = 10
UserID = 1
recom <- predict(UBCF_model, ratingmat[UserID], n=items_to_recommend) 
recom_list <- as(recom, "list") #convert recommenderlab object to readable list

# Obtain recommendations
for (i in c(1:items_to_recommend)){
 recom_result[i] <- movies[as.integer(recom_list[[1]][i]),2]
}
recom_result
```

## 2. Item-based recommendation:

Item-based recommendation is a form of collaborative filtering for recommender systems that uses the similarity between items which calculated using people's ratings of those items. 

- **Build Recommender Model**

Build Recommender Model using Item-Based Collaborative Filtering using `recommenderlab` library. Set method equals to "IBCF", use Cosine Similarity as parameters.
```{r}
# Create Recommender Model using Item-Based Collaborative Filtering
IBCF_model <- Recommender(getData(eval_sets, "train"), "IBCF", 
                           param=list(normalize = "Z-score",method="Cosine",k=30))
```

```{r echo = FALSE, message = FALSE, warning = FALSE}
# Save the model
saveRDS(IBCF_model, file = "data/IBCF_recommender.rds")
# UBCF_model <- readRDS("data/IBCF_recommender.rds")
```

- **Evaluate your model:**

To evaluate the model we use the `recommenderlab` package. We first use the `predict()` function to generate predictions for the known portion of the test data. Then, the `calcPredictAccuracy()` function to calculate the error between the predictions and the unknown portions of the test data.

```{r}
items_to_recommend = 10
IBCF_eval_prediction = predict(object = IBCF_model,
                          newdata = getData(eval_sets, "known"),
                          n = items_to_recommend,
                          type = "ratings")
# Get model performance
IBCF_eval_accuracy = calcPredictionAccuracy(x = IBCF_eval_prediction,
                                       data = getData(eval_sets, "unknown"),
                                       byUser = FALSE)
IBCF_eval_accuracy
```
- **Get recommendations:**

```{r}
# Obtain top 10 recommendations for n user in dataset
UserID = 15
recom <- predict(IBCF_model, ratingmat[UserID], n=items_to_recommend) 
recom_list <- as(recom, "list") #convert recommenderlab object to readable list
 
# Obtain recommendations
recom_result <- matrix(0,items_to_recommend)
for (i in c(1:items_to_recommend)){
 recom_result[i] <- movies[as.integer(recom_list[[1]][i]),2]
}
recom_result
```

## 3. Report User-based vs. Item-based accuracy:

Both models use Cosine Similarity, Z-score normalization and 30 nearest neighbors or items as parameters.

```{r}
# Aggregate the performance statistics
performance <- rbind(
 UBCF_eval_accuracy,
 IBCF_eval_accuracy
)
kable(performance)
```

User-based collaborative model (UBCF) outperforms Item-based collaborative model (IBCF). On the other hand, UBCF has greater computational cost and requires more resources. 

***


