---
title: "Project 2: Movie Recommendation"
author: 
  - Tatiana Sokolinski, tatiana9@illinois.edu
  - Noam Mansovsky, noamm2@illinois.edu
date: "12/11/2021"
output:
  html_document: 
    theme: yeti
    toc: yes
urlcolor: cyan
---
***
```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen = 1, digits = 4, width = 80, fig.align = "center")

mypackages = c( "dplyr","ggplot2","recommenderlab","DT","data.table","reshape2","Matrix","proxy","doBy")   
tmp = setdiff(mypackages, rownames(installed.packages())) 
if (length(tmp) > 0) install.packages(tmp)

library(dplyr)
library(ggplot2)
library(recommenderlab)
library(DT)
library(data.table)
library(reshape2)
library(Matrix)
library(proxy)
library(doBy) # to get top-n elements
library(knitr)
```


## Read in Data

### Ratings data:
```{r}
myurl = "https://liangfgithub.github.io/MovieData/"
ratings = read.csv(paste0(myurl, 'ratings.dat?raw=true'), 
                   sep = ':',
                   colClasses = c('integer', 'NULL'), 
                   header = FALSE)
colnames(ratings) = c('UserID', 'MovieID', 'Rating', 'Timestamp')
head(ratings, 5)
```

### User data:
```{r}
users = read.csv(paste0(myurl, 'users.dat?raw=true'),
                 sep = ':', header = FALSE)
users = users[, -c(2,4,6,8)] # skip columns
colnames(users) = c('UserID', 'Gender', 'Age', 'Occupation', 'Zip-code')
head(users, 5)
```

### Movies data:

```{r}
movies = readLines(paste0(myurl, 'movies.dat?raw=true'))
movies = strsplit(movies, split = "::", fixed = TRUE, useBytes = TRUE)
movies = matrix(unlist(movies), ncol = 3, byrow = TRUE)
movies = data.frame(movies, stringsAsFactors = FALSE)
colnames(movies) = c('MovieID', 'Title', 'Genres')
movies$MovieID = as.integer(movies$MovieID)

# convert accented characters
movies$Title[73]
movies$Title = iconv(movies$Title, "latin1", "UTF-8")
movies$Title[73]

# extract year
movies$Year = as.numeric(unlist(
  lapply(movies$Title, function(x) substr(x, nchar(x)-4, nchar(x)-1))))
head(movies, 5)
```

***
# System I: Content-Based Recommendation System

## 1. Recommend the top-five "highly-rated" movies in the selected genre:

**Examine the data:**

Number of moview in each rating category:
```{r}
# Count the number of movies in each rating category:
ratings %>% group_by(Rating) %>% summarize(n=n())
```

Average number of time movie receives the rating:
```{r}
# Count the number of time movie receives the rating:
rating_cnt <- ratings %>% group_by(MovieID) %>% summarize(n=n())
head(rating_cnt)
```
The plot below shows distribution of number of movies ratings. On average movie received about `r mean(rating_cnt$n, na.rm=T)` users reviews.
```{r}
ggplot(rating_cnt, aes(x=n)) +
    geom_histogram(colour="DodgerBlue", fill="white",bins=50) +
    geom_vline(aes(xintercept=mean(n, na.rm=T)),  
               color="orange", linetype="dashed", size=1)+
  ggtitle("Distribution Movies Ratings") +
  geom_text(aes(x=mean(n, na.rm=T), label=round(mean(n, na.rm=T)), y=300), colour="DimGrey", angle=90) +
  xlab("Count") + ylab("Number of Review")
```

- **Create a table that summarizes all genres for each movie:**

```{r}
# Subset Genres column
genres <- as.data.frame(movies$Genres, stringsAsFactors=FALSE)
# Split text on pipe to get all genres for each movie
tmp = as.data.frame(tstrsplit(genres[,1], '[|]',
                              type.convert=TRUE),
                    stringsAsFactors=FALSE)
colnames(tmp) <- c(1:dim(tmp)[2])
head(tmp)
```
- **Create the movie feature matrix with binary indicator for each genre:**

```{r}
genre_list = c("Action", "Adventure", "Animation", 
               "Children's", "Comedy", "Crime",
               "Documentary", "Drama", "Fantasy",
               "Film-Noir", "Horror", "Musical", 
               "Mystery", "Romance", "Sci-Fi", 
               "Thriller", "War", "Western")

genre_matrix = matrix(0, nrow(movies), length(genre_list))
# Loop over all movies
for(i in 1:nrow(tmp)){
  genre_matrix[i,genre_list %in% tmp[i,]]=1
}
colnames(genre_matrix) = genre_list
head(genre_matrix)
```

- **Create the user profile matrix:**

Transform the rating dataset from a long format to a wide format. Here, we can see that UserID 8 has watched MovieID 1 & 4 and gave them 4 & 3 points. We can also say that he did not rated MovieID 2,3,5 & 6, therefore they have rating 0.
```{r}
ratings = ratings[, -c(4)]
ratings_long <- dcast(ratings, MovieID~UserID, value.var = "Rating", na.rm=FALSE)
# for (i in 1:ncol(ratings_long)){
#   ratings_long[which(is.na(ratings_long[,i]) == TRUE),i] <- 0
# }
ratings_long <- ratings_long[,-1] #remove movieIds col. Rows are movieIds, cols are userIds
# Remove movies with no reviews
ratings_long <- ratings_long[rowSums(is.na(ratings_long)) != ncol(ratings_long), ]
ratings_long[1:6, 1:9]
```


- **Recommend top-five highly-rated movies in the selected genre:**

The methodology to find the top-five **"highly-rated"** movies. For each movie we calculate movie weighted rating score:

Movie Weighted Rating $(MWR) = (\frac{v}{v + m}) * R + (\frac{m}{v + m})*C$

Where:

- R - average rating for the movie

- v - count of reviews for the movie

- m - minimum review needed to be considered for "highly-rated" category. Since on average movie receives 270 users reviews, we can consider only movies with 50 or more reviews. m = 50

- C - constant that represent the average rating for the movie in this genre.


```{r}
# Select genre
genre <- genre_list[18]
genre
# List of movies of this genre
movie_list <- which(genre_matrix[,c(genre)] == 1)
movies_options <- ratings_long[movie_list, ]

m = 50
# Average movie review
R <- rowMeans(movies_options, na.rm = TRUE)
# Number of reviews received
v <- rowSums(!is.na(movies_options))
# Average rating for the movie in this genre
C <- mean(R, na.rm = TRUE)
# Compute Movie Weighted Rating
movies_options['MWR'] = (v / (v + m)) * R + (m / (m +v)) * C
movies_options['indx'] <- movie_list
movies_options['ave_rating'] <- R
movies_options['n_reviews'] <- v

```

Recommend top-five "highly-rated" movies:
```{r}
# Select the top-five "highly-rated" movies in the specified genre
table <- movies_options[movies_options$n_reviews >=50, c('indx','n_reviews','MWR')]
rows <- table[which.maxn(table$MWR, n=5) , 'indx']
# Recommend movies
movies[rows,]
```

## 2. Recommend the top-five "most-popular" movies in the selected genre:

- **Create the user profile matrix:**

This time the user profile matrix uses a binary indicator "1" when a movie rating equals 4 or 5 and "0" when a movie rating equals 1,2 or 3.
```{r}
ratings_binary <- ratings
ratings_binary$Rating[ratings_binary$Rating == 1 | ratings_binary$Rating == 2 | ratings_binary$Rating == 3 ] <- 0   
ratings_binary$Rating[ratings_binary$Rating == 4 | ratings_binary$Rating == 5 ] <- 1 
ratings_binary = ratings_binary[, -c(4)]
head(ratings_binary)
```
Transform the binary rating dataset from a long format to a wide format:
```{r}
ratings_binary_long <- dcast(ratings_binary, MovieID~UserID, value.var = "Rating", na.rm=FALSE)
ratings_binary_long = ratings_binary_long[,-1] #remove movieIds col. Rows are movieIds, cols are userIds
head(ratings_binary_long[1:6,1:6])
```

- **Recommend top-five most-popular movies in the selected genre:**

The methodology to find the top-five **"most-popular"** movies. For each movie we calculate movie popularity score:

Movie Popularity Score $(MPS) = P + P * \frac{P}{N}$

Where:

- P - count of positive reviews (where user's review >= 4)

- N - count of negative reviews (where user's review <= 3)


```{r}
# Select genre
genre <- genre_list[18]
genre
# List of movies of this genre
movie_list <- which(genre_matrix[,c(genre)] == 1)
movies_options <- ratings_binary_long[movie_list, ]


# Number of reviews received
v <- rowSums(!is.na(movies_options)) 

# Total positive and negative reviews
positive_reviews <- rowSums(movies_options, na.rm = TRUE)
negative_reviews <- v - positive_reviews

movies_options['indx'] <- movie_list
movies_options['n_reviews'] <- v
movies_options['pos_reviews'] <- positive_reviews
movies_options['neg_reviews'] <- negative_reviews
movies_options['ratio'] <- positive_reviews / negative_reviews
movies_options['MPS'] <- positive_reviews + positive_reviews * movies_options['ratio']
data <- movies_options[, c('indx','n_reviews','pos_reviews','neg_reviews','ratio','MPS')]
data <- data %>% mutate(MPS = coalesce(MPS, 0)) # fill na with zeros
data <- data %>% mutate_if(is.numeric, function(x) ifelse(is.infinite(x), 0, x)) # fill na with zeros
```

Recommend top-five "most-popular" movies:
```{r}
# Select the top-five "most-popular" movies in the specified genre
rows <- table[which.maxn(data$MPS, n=5) , 'indx']
# Recommend movies
movies[rows,]
```


***
# System II: Collaborative Recommendation System

## 1. User-based recommendation:

User-Based Collaborative Filtering is an algorithm used to predict the items that a user might like on the basis of ratings given by the other users who have similar taste with that of the target user.

- **Create the user profile matrix:**

Create ratings matrix using MovieID, UserID and Ratings fields.
```{r}
# Create ratings matrix. Rows = UserID, Columns = MovieID
ratingmat <- dcast(ratings, UserID~MovieID, value.var = "Rating", na.rm=FALSE)
ratingmat <- as.matrix(ratingmat[,-1]) #remove userIds
# Convert rating matrix into a recommenderlab sparse matrix
ratingmat <- as(ratingmat, "realRatingMatrix")
dim(ratingmat)
```

- **Train and test splits:**

The `evaluationScheme()` function can split the data set into training and testing subsets:

- **train** matrix contains about 80% rows of `rating` dataset

- **test** matrix contains about 20% rows of `rating` dataset


The output of the evaluationScheme() function is a single R object containing both the training and testing subsets.

```{r}
set.seed(9335)
# Parameters:
percent_train = 0.8 # split the data into the training and the test sets (80%/20%)
items_to_keep = 15  # items to use for each user
rating_threshold = 4 # threshold at which ratings are considered good for evaluation
eval_sets <- evaluationScheme(ratingmat, method="split", train=percent_train, given=items_to_keep, goodRating=rating_threshold)
eval_sets
```

- **Build Recommender Model**

Build Recommender Model using User-Based Collaborative Filtering using `recommenderlab` library. Set method equals to "UBCF", use Cosine Similarity, Z-score normalization and 30 Nearest Neighbors as parameters.
```{r}
# Create Recommender Model using User-Based Collaborative Filtering
UBCF_model <- Recommender(getData(eval_sets, "train"), "UBCF", 
                           param=list(normalize = "Z-score",method="Cosine",nn=30))
```

- **Evaluate your model:**

To evaluate the model we use the `recommenderlab` package. We first use the `predict()` function to generate predictions for the known portion of the test data. Then, the `calcPredictAccuracy()` function to calculate the error between the predictions and the unknown portions of the test data.

```{r}
items_to_recommend = 10
UBCF_eval_prediction = predict(object = UBCF_model,
                          newdata = getData(eval_sets, "known"),
                          n = items_to_recommend,
                          type = "ratings")
# Get model performance
UBCF_eval_accuracy = calcPredictionAccuracy(x = UBCF_eval_prediction,
                                       data = getData(eval_sets, "unknown"),
                                       byUser = FALSE)
UBCF_eval_accuracy
```
- **Get recommendations:**

```{r}
# Obtain top 10 recommendations for n user in dataset
UserID = 1
recom <- predict(UBCF_model, ratingmat[UserID], n=items_to_recommend) 
recom_list <- as(recom, "list") #convert recommenderlab object to readable list
 
# Obtain recommendations
recom_result <- matrix(0,items_to_recommend)
for (i in c(1:items_to_recommend)){
 recom_result[i] <- movies[as.integer(recom_list[[1]][i]),2]
}
recom_result
```

## 2. Item-based recommendation:

Item-based recommendation is a form of collaborative filtering for recommender systems that uses the similarity between items which calculated using people's ratings of those items. 

- **Build Recommender Model**

Build Recommender Model using Item-Based Collaborative Filtering using `recommenderlab` library. Set method equals to "IBCF", use Cosine Similarity as parameters.
```{r}
# Create Recommender Model using Item-Based Collaborative Filtering
IBCF_model <- Recommender(getData(eval_sets, "train"), "IBCF", 
                           param=list(normalize = "Z-score",method="Cosine",k=30))
```

- **Evaluate your model:**

To evaluate the model we use the `recommenderlab` package. We first use the `predict()` function to generate predictions for the known portion of the test data. Then, the `calcPredictAccuracy()` function to calculate the error between the predictions and the unknown portions of the test data.

```{r}
items_to_recommend = 10
IBCF_eval_prediction = predict(object = IBCF_model,
                          newdata = getData(eval_sets, "known"),
                          n = items_to_recommend,
                          type = "ratings")
# Get model performance
IBCF_eval_accuracy = calcPredictionAccuracy(x = IBCF_eval_prediction,
                                       data = getData(eval_sets, "unknown"),
                                       byUser = FALSE)
IBCF_eval_accuracy
```
- **Get recommendations:**

```{r}
# Obtain top 10 recommendations for n user in dataset
UserID = 15
recom <- predict(IBCF_model, ratingmat[UserID], n=items_to_recommend) 
recom_list <- as(recom, "list") #convert recommenderlab object to readable list
 
# Obtain recommendations
recom_result <- matrix(0,items_to_recommend)
for (i in c(1:items_to_recommend)){
 recom_result[i] <- movies[as.integer(recom_list[[1]][i]),2]
}
recom_result
```

## 3. Report User-based vs. Item-based accuracy:

Both models use Cosine Similarity, Z-score normalization and 30 nearest neighbors or items as parameters.

```{r}
# Aggregate the performance statistics
performance <- rbind(
 UBCF_eval_accuracy,
 IBCF_eval_accuracy
)
kable(performance)
```

User-based collaborative model (UBCF) outperforms Item-based collaborative model (IBCF). On the other hand, UBCF has greater computational cost and requires more resources. 


***
