---
title: "Movielense Data: Recommendation Based on Genres"
author: 
  - Tatiana Sokolinski, tatiana9@illinois.edu
  - Noam Mansovsky, noamm2@illinois.edu
date: "12/2/2021"
output:
  html_document: 
    theme: yeti
    toc: yes
urlcolor: cyan
---
***
```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(scipen = 1, digits = 4, width = 80, fig.align = "center")

mypackages = c( "dplyr","ggplot2","recommenderlab","DT","data.table","reshape2","Matrix","proxy","doBy")   
tmp = setdiff(mypackages, rownames(installed.packages())) 
if (length(tmp) > 0) install.packages(tmp)

library(dplyr)
library(ggplot2)
library(recommenderlab)
library(DT)
library(data.table)
library(reshape2)
library(Matrix)
library(proxy)
library(doBy) # to get top-n elements
```


## Read in Data

### Ratings data:
```{r}
myurl = "https://liangfgithub.github.io/MovieData/"
ratings = read.csv(paste0(myurl, 'ratings.dat?raw=true'), 
                   sep = ':',
                   colClasses = c('integer', 'NULL'), 
                   header = FALSE)
colnames(ratings) = c('UserID', 'MovieID', 'Rating', 'Timestamp')
head(ratings, 5)
```

### User data:
```{r}
users = read.csv(paste0(myurl, 'users.dat?raw=true'),
                 sep = ':', header = FALSE)
users = users[, -c(2,4,6,8)] # skip columns
colnames(users) = c('UserID', 'Gender', 'Age', 'Occupation', 'Zip-code')
head(users, 5)
```

### Movies data:

```{r}
movies = readLines(paste0(myurl, 'movies.dat?raw=true'))
movies = strsplit(movies, split = "::", fixed = TRUE, useBytes = TRUE)
movies = matrix(unlist(movies), ncol = 3, byrow = TRUE)
movies = data.frame(movies, stringsAsFactors = FALSE)
colnames(movies) = c('MovieID', 'Title', 'Genres')
movies$MovieID = as.integer(movies$MovieID)

# convert accented characters
movies$Title[73]
movies$Title = iconv(movies$Title, "latin1", "UTF-8")
movies$Title[73]

# extract year
movies$Year = as.numeric(unlist(
  lapply(movies$Title, function(x) substr(x, nchar(x)-4, nchar(x)-1))))
head(movies, 5)
```

***
# System I: Content-Based Recommendation System

## 1. Recommend the top-five "highly-rated" movies in the selected genre:

**Examine the data:**

Number of moview in each rating category:
```{r}
# Count the number of movies in each rating category:
ratings %>% group_by(Rating) %>% summarize(n=n())
```

Average number of time movie receives the rating:
```{r}
# Count the number of time movie receives the rating:
rating_cnt <- ratings %>% group_by(MovieID) %>% summarize(n=n())
head(rating_cnt)
```
The plot below shows distribution of number of movies ratings. On average movie received about `r mean(rating_cnt$n, na.rm=T)` users reviews.
```{r}
ggplot(rating_cnt, aes(x=n)) +
    geom_histogram(colour="DodgerBlue", fill="white",bins=50) +
    geom_vline(aes(xintercept=mean(n, na.rm=T)),  
               color="orange", linetype="dashed", size=1)+
  ggtitle("Distribution Movies Ratings") +
  geom_text(aes(x=mean(n, na.rm=T), label=round(mean(n, na.rm=T)), y=300), colour="DimGrey", angle=90) +
  xlab("Count") + ylab("Number of Review")
```

- **Create a table that summarizes all genres for each movie:**

```{r}
# Subset Genres column
genres <- as.data.frame(movies$Genres, stringsAsFactors=FALSE)
# Split text on pipe to get all genres for each movie
tmp = as.data.frame(tstrsplit(genres[,1], '[|]',
                              type.convert=TRUE),
                    stringsAsFactors=FALSE)
colnames(tmp) <- c(1:dim(tmp)[2])
head(tmp)
```
- **Create the movie feature matrix with binary indicator for each genre:**

```{r}
genre_list = c("Action", "Adventure", "Animation", 
               "Children's", "Comedy", "Crime",
               "Documentary", "Drama", "Fantasy",
               "Film-Noir", "Horror", "Musical", 
               "Mystery", "Romance", "Sci-Fi", 
               "Thriller", "War", "Western")

genre_matrix = matrix(0, nrow(movies), length(genre_list))
# Loop over all movies
for(i in 1:nrow(tmp)){
  genre_matrix[i,genre_list %in% tmp[i,]]=1
}
colnames(genre_matrix) = genre_list
head(genre_matrix)
```

- **Create the user profile matrix:**

Transform the rating dataset from a long format to a wide format. Here, we can see that UserID 8 has watched MovieID 1 & 4 and gave them 4 & 3 points. We can also say that he did not rated MovieID 2,3,5 & 6, therefore they have rating 0.
```{r}
ratings = ratings[, -c(4)]
ratings_long <- dcast(ratings, MovieID~UserID, value.var = "Rating", na.rm=FALSE)
# for (i in 1:ncol(ratings_long)){
#   ratings_long[which(is.na(ratings_long[,i]) == TRUE),i] <- 0
# }
ratings_long <- ratings_long[,-1] #remove movieIds col. Rows are movieIds, cols are userIds
# Remove movies with no reviews
ratings_long <- ratings_long[rowSums(is.na(ratings_long)) != ncol(ratings_long), ]
ratings_long[1:6, 1:9]
```


- **Recommend top-five highly-rated movies in the selected genre:**

The methodology to find the top-five **"highly-rated"** movies. For each movie we calculate movie weighted rating score:

Movie Weighted Rating $(MWR) = (\frac{v}{v + m}) * R + (\frac{m}{v + m})*C$

Where:

- R - average rating for the movie

- v - count of reviews for the movie

- m - minimum review needed to be considered for "highly-rated" category. Since on average movie receives 270 users reviews, we can consider only movies with 50 or more reviews. m = 50

- C - constant that represent the average rating for the movie in this genre.


```{r}
# Select genre
genre <- genre_list[18]
genre
# List of movies of this genre
movie_list <- which(genre_matrix[,c(genre)] == 1)
movies_options <- ratings_long[movie_list, ]

m = 50
# Average movie review
R <- rowMeans(movies_options, na.rm = TRUE)
# Number of reviews received
v <- rowSums(!is.na(movies_options))
# Average rating for the movie in this genre
C <- mean(R, na.rm = TRUE)
# Compute Movie Weighted Rating
movies_options['MWR'] = (v / (v + m)) * R + (m / (m +v)) * C
movies_options['indx'] <- movie_list
movies_options['ave_rating'] <- R
movies_options['n_reviews'] <- v

```

Recommend top-five "highly-rated" movies:
```{r}
# Select the top-five "highly-rated" movies in the specified genre
table <- movies_options[movies_options$n_reviews >=50, c('indx','n_reviews','MWR')]
rows <- table[which.maxn(table$MWR, n=5) , 'indx']
# Recommend movies
movies[rows,]
```



***
